\texttt{apt-p2p}: A Peer-to-Peer Distributor for Free Software Package Release and Update

### 2008-08-26 docs/paper/paper.tex
Added JC's changes to the paper.

\title{Peer-to-Peer Distribution of Free Software Packages}
\author{\IEEEauthorblockN{Cameron Dale}
\IEEEauthorblockA{School of Computing Science\\
Simon Fraser University\\
Burnaby, British Columbia, Canada\\
Email: camerond@cs.sfu.ca}
\and
\IEEEauthorblockN{Jiangchuan Liu}
\IEEEauthorblockA{School of Computing Science\\
Simon Fraser University\\
Burnaby, British Columbia, Canada\\
Email: jcliu@cs.sfu.ca}}

@@

This paper is organized as follows. The current situation and its
problems are presented in section \ref{situation}. We analyze some
related solutions in section \ref{related}, and then present our
solution in section \ref{opportunity}. We then detail our sample
implementation for Debian-based distributions in section
\ref{implementation}, including an in-depth look at our DHT
customizations in section \ref{custom_dht}. We analyze our deployed
sample implementation in section \ref{analysis}, suggest some future
enhancements in section \ref{future}, and conclude the paper in
section \ref{conclusions}.

@@

\section{Problem/Situation}
\label{situation}

There are a large number of groups using the Internet to distribute
their free software packages. These packages are usually available
from a free download site, which usually requires a network of
mirrors to support the large number of requests. These distribution
systems almost always support some kind of file verification,
usually cryptographic hashes, to verify the completed downloads
accuracy or authenticity.

\subsection{Examples}
\label{examples}

@@

Other software distributors also use a similar system. CPAN
distributes software packages for the PERL
programming language, using SOAP RPC requests to find and download
files. Cygwin provides many of the
standard Unix/Linux tools in a Windows environment, using a
package management tool that requests packages from websites. There
are two software distribution systems for Mac OSX, fink and
MacPorts, that also retrieve packages in this way.

Also, some systems use direct web downloads, but with a hash
verification file also available for download next to the desired
file. These hash files usually have the same file name, but with an
added extension identifying the hash used (e.g. \texttt{.md5} for
the MD5 hash). This type of file downloading and verification is
typical of free software hosting facilities that are open to anyone
to use, such as SourceForge.

\subsection{Similarities}
\label{similarities}

The important things to note for each of the systems mentioned in
section \ref{examples}, is that they all have the following in
common:
\begin{itemize}
 \item The content is divided into distinct units (packages), each
       of which contains a small independent part of all the
       content.
 \item The packages are avaliable for anyone to freely download.
 \item Users are typically not interested in downloading all of the
       packages available.
 \item Hashes of the packages are available before the download is
       attempted.
 \item Requests to download the packages are sent by a tool, not
       directly by the user (though the tool is responding to
       requests from the user).
\end{itemize}

We also expect that there are a number of users of these systems
that are motivated by altruism to want to help out with this
distribution. This is common in these systems due to the free nature
of the software being delivered, which encourages some to want to
help out in some way. A number of the systems are also used by
groups that are staffed mostly, or sometimes completely, by
volunteers. This also encourages users to want to give back
to the volunteer community that has created the software they are
using.

\subsection{Differences}
\label{differences}

Although at first it might seem that a single all-reaching solution
is possible for this situation, there are some differences in each
system that require independent solutions. The systems all use
different tools for their distribution, so any implementation would
have to be specific to the tool it is to integrate with. In
particular, how to obtain requests from the tool or user, and how to
find a hash that corresponds to the file being requested, is very
specific to each system.

Also, there may be no benefit in creating a single large solution to
integrate all these problems. Although they sometimes distribute
nearly identical packages (e.g. the same software available in
multiple Linux distributions), it is not exactly identical and has
usually been tailored to the system by the distributor. The small
differences will change the hash of the files, and so will make it
impossible to distribute similar packages across systems. And,
although peer-to-peer systems can scale very well with the number of
peers in the system, there is some overhead involved, so having a
much larger system of peers would mean that requests could take
longer to complete.

\subsection{Problems}
\label{problems}

There are some attributes of software package distribution that make
it difficult to use an existing peer-to-peer solution with, or to
design a new solution for this need.

@@

The dimensions of the software archive makes it difficult to
distribute efficiently. While most of the packages are very small in
size, there are some that are quite large. There are too many
packages to distribute each individually, but the archive is also
too large to distribute in its entirety. In some archives there are
also divisions of the archive into sections, e.g. by the computer
architecture that the package is intended for.

@@

For example, Figure~\ref{size_CDF} shows the size of packages in the
Debian distribution. While 80\% of the packages are less than
512~KB, some of the packages are hundreds of megabytes. The entire
archive consists of 22,298 packages and is approximately 119,000 MB
in size. Some packages are divided by architecture, but there are a
large number of packages that are not architecture specific and so
can be installed on any computer architecture.

@@

The software packages being distributed are being constantly
updated. These updates could be the result of the software creators
releasing a new version of the software with improved functionality,
or the result of the distributor updating their packaging of the
software to meet new requirements. Even if the distributor
periodically makes \emph{stable} releases, which are snapshots of
all the packages in the archive at a certain time, updates are still
released for security issues or serious bugs.

@@

For example, Figure~\ref{update_size} shows the amount of data in
the Debian archive that was updated each day over a period of 3
months. Each day approximately 1.5\% of the 119,000 MB archive is
updated with new versions of packages.

@@

Though there are a large number of packages, and a large number of
users, interest in a given version of a package could be very
limited. If the package is necessary for the system (e.g. the
package distribution software), then it is likely that everyone will
have it installed, but there are very few of these packages. Most
packages fall in the category of optional or extra, and so are
interesting to only a limited number of people.

@@

\subsubsection{Interactive Users}

The package management software that downloads packages displays
some kind of indication of speed and completeness for users to
watch. Since previous client-server downloads occur in a sequential
fashion, the package management software also measures the speed
based on sequential downloading. This requires the peer-to-peer
solution to be reasonably responsive at retrieving packages
sequentially.

@@

\section{Related Work/Existing Solutions}
\label{related}

There are many peer-to-peer implementations available today, but
none is very well suited to this specific problem.

\subsection{BitTorrent}
\label{bittorrent}

Many distributors make their software available in some form using
BitTorrent \cite{COHEN03}, e.g. for the distribution of CD
images. This is not an ideal situation though, as it requires the
peers to download large numbers of packages that they are not
interested in, and prevents them from updating to new packages
without downloading another image containing a lot of the same
packages they already have. Unfortunately, BitTorrent is not ideally
suited to an application such as this one.

First, there is no obvious way to divide the packages into torrents.
Most of the packages are too small, and there are too many packages
in the entire archive, to create individual torrents for each one.
However, all the packages together are too large to track
efficiently as a single torrent. Some division of the archive's
packages into torrents is obviously necessary, but wherever that
split occurs it will cause either some duplication of connections,
or prevent some peers from connecting to others who do have the same
content. Also, a small number of the packages can be updated every
day which would add new files to the torrent, thereby changing its
\emph{infohash} identifier and making it a new torrent. This will
severely fracture the download population, even though peers in the
new torrent share 99\% of the packages in common with peers in the
old torrent.

Other issues also prevent BitTorrent from being a good solution to
this problem. BitTorrent's fixed piece sizes that disregard file
boundaries are bigger than many of the packages in the archive. This
will waste peers' downloading bandwidth as they will end up
downloading parts of other packages just to get the piece that
contains the package they do want. Incentives to share (upload) are
no longer needed, as the software is freely available for anyone to
download without sharing. Seeds are also not needed as the mirrors
can serve in that capacity. Finally, BitTorrent downloads files
randomly, which does not work well with the interactive package
management tools expectation of sequential downloads.

\subsection{Other Solutions}
\label{others}

There have also been other attempted implementations, usually based
on some form of BitTorrent, to address this problem. Some other
implementations have used BitTorrent almost unmodified, while others
have only looked at using DHTs to replace the tracker in a
BitTorrent system. apt-torrent \cite{apttorrent} creates torrents
for some of the larger packages available, but it ignores the
smaller packages, which are often the most popular. DebTorrent
\cite{debtorrent} makes widespread modifications to a traditional
BitTorrent client, to try and fix the drawbacks mentioned in section
\ref{bittorrent}. However, these changes also require some
modifications to the distribution system to support it.

@@

\section{Opportunity/Our Proposal}
\label{opportunity}

@@

\section{Sample Implementation/Implementation}
\label{implementation}

@@

\section{Analysis}
\label{analysis}

@@

@@

\section{Future Work}
\label{future}

@@

\section{Conclusions}
\label{conclusions}

>>>>

\title{\texttt{apt-p2p}: A Peer-to-Peer Distributor for Free Software Package Release and Update}
\author{\IEEEauthorblockN{Cameron Dale}
\IEEEauthorblockA{School of Computing Science\\
Simon Fraser University\\
Burnaby, British Columbia, Canada\\
Email: camerond@cs.sfu.ca}
\and
\IEEEauthorblockN{Jiangchuan Liu}
\IEEEauthorblockA{School of Computing Science\\
Simon Fraser University\\
Burnaby, British Columbia, Canada\\
Email: jcliu@cs.sfu.ca}}

@@

The rest of this paper is organized as follows. The background and motivation are presented in Section \ref{situation}, together with related works in  Section \ref{related}. We propose
our solution in Section \ref{opportunity}. We then detail our sample
implementation for Debian-based distributions in Section
\ref{implementation}, including an in-depth look at our DHT
customizations in Section \ref{custom_dht}. Its performance is evaluated in Section \ref{analysis}. Finally, 
Section \ref{conclusions} concludes the paper and offers some future directions.

@@

\section{Background and Motivations}
\label{situation}

In the free software society, there are a large number of groups using the Internet to 
collaboratively develop and release their software. The ever increasing power of
modern programming languages and operating systems has made these software, like commercial software, extremely large and complex, which often
consists of many small units (packages). Together with their popularity among users, 
an efficient and reliable management and distribution of these packages over the Internet has become an daunting task. In this section, we offer concrete examples illustrating the 
unique challenges in this context.

\subsection{Free Software Package Distributor: Examples}
\label{examples}

@@

Similar tools have been used for other software packages. CPAN
distributes software packages for the PERL
programming language, using SOAP RPC requests to find and download
files. Cygwin provides many of the
standard Unix/Linux tools in a Windows environment, using a
package management tool that requests packages from websites. There
are two software distribution systems for Mac OSX, fink and
MacPorts, that also retrieve packages in this way.

Direct web downloading is also a common way, often coupled with a hash
verification file to be downloaded next to the desired
file. The hash file usually have the same file name, but with an
added extension identifying the hash used (e.g. \texttt{.md5} for
the MD5 hash). This type of file downloading and verification is
typical of free software hosting facilities that are open to anyone
to use, such as SourceForge.

Given the free nature of these software, there are often a number of users 
motivated by altruism to want to help out with their distribution. 
This is particularly true considering that many of these software are used by
groups that are staffed mostly, or sometimes completely, by
volunteers. They are thus motivated to contribute their network resources, so as to promote the healthy development 
of the volunteer community that released the software. As a matter of fact,
we have see many free mirror sites hosting these software packages for downloading. 
We also naturally expect that peer-to-peer distribution can be implementation in 
this context, which scale well with large user bases and can easily explore the network resources made available by
the volunteers.

\subsection{Unique Characteristics}
\label{problems}

While it seems straightforward to use an existing peer-to-peer file sharing tool like BitTorrent for
free software package distribution, there are indeed a series of new challenges in this unique scenario:

@@

While most of the packages of a software release are very small in
size, there are some that are quite large. There are too many
packages to distribute each individually, but the archive is also
too large to distribute in its entirety. In some archives there are
also divisions of the archive into sections, e.g. by the operating system (OS) or computer
architecture that the package is intended for.

@@

For example, Figure~\ref{size_CDF} shows the size of the packages in the
current Debian distribution. While 80\% of the packages are less than
512~KB, some of the packages are hundreds of megabytes. The entire
archive consists of 22,298 packages and is approximately 119,000 MB
in size. Most of the packages are to be installed in any computer environment, but there are 
also OS- or architecture-specific packages, as shown by the adjusted sizes based on popularity of the packages. ((((more words on how the size is adjusted by popularity))))

@@

The software packages being distributed are being constantly
updated. These updates could be the result of the software creators
releasing a new version with improved functionality,
or the distributor updating their packaging of the
software to meet new requirements. Even if the distributor
periodically makes \emph{stable} releases, which are snapshots of
all the packages in the archive at a certain time, updates are still
released for security issues or serious bugs.

@@

For example, Figure~\ref{update_size} shows the amount of data in
the Debian archive that was updated each day over a period of 3
months. In every single day, approximately 1.5\% of the 119,000 MB archive is
updated with new versions of packages. Note that this frequency is much higher than
that of most commercial software, mainly because many free software are
developed in a loosely management environment with developers working
asynchronously from worldwide.

@@

Finally, though there are a large number of packages and a large number of
users, the interest in a particular version of a package can be very 
limited. Specifically, there are core packages that every user has to download, but most
packages would fall in the category of optional or extra, and so are
interesting to only a limited number of people.

@@

@@

\subsection{Why BitTorrent Doesn't Work Well}
\label{related}

Recently, many distributors make their software available using
BitTorrent \cite{COHEN03}, in particular, for the distribution of CD
images. This straightforward use however is far ineffective, as it requires the
peers to download large numbers of packages that they are not
interested in, and prevents them from updating to new packages
without downloading another image containing a lot of the same
packages they already have.

An alternative is to create torrents tracking individual packages. Unfortunately, we find that this enhancement can be
quite difficult given the unique characteristic of free software packages.

First, there is no obvious way to divide the packages into torrents.
Most of the packages are too small, and there are too many packages
in the entire archive to create individual torrents for each one.
On the other hand, all the packages together are too large to track
efficiently as a single torrent. Hence, some division of the archive's
packages into torrents is obviously necessary, but wherever that
split occurs it will cause either some duplication of connections,
or prevent some peers from connecting to others who do have the same
content. In addition, a small number of the packages can be updated every
day which would add new files to the torrent, thereby changing its
\emph{infohash} identifier and making it a new torrent. This will
severely fracture the download population, even though peers in the
new torrent may share 99\% of the packages in common with peers in the
old torrent.

Other issues also prevent BitTorrent from being a good solution to
this problem. In particular, BitTorrent's fixed piece sizes (?KB) that disregard file
boundaries are bigger than many of the packages in the archive. This
will waste peers' downloading bandwidth as they will end up
downloading parts of other packages just to get the piece that
contains the package they do want.

Finally, note that BitTorrent downloads files
randomly, which does not work well with the interactive package
management tools expectation of sequential downloads. On the other hand, with altruistic peers, incentives to share (upload) 
become a less important issue, and the availability of seeds are not critical, either, as the mirror sites 
can serve in that capacity.

@@

\section{Peer-to-Peer Assisted Distributor: An Overview}
\label{opportunity}

@@

\section{\texttt{apt-p2p}: A Practical Implementation}
\label{implementation}

@@

\section{Performance Evaluation}
\label{analysis}

@@

\section{Related Work}
\label{others}

There have also been preliminary attempts to implement peer-to-peer distributors for
software packages. apt-torrent \cite{apttorrent} creates torrents
for some of the larger packages available, but it ignores the
smaller packages, which are often the most popular. DebTorrent
\cite{debtorrent} makes widespread modifications to a traditional
BitTorrent client, to try and fix the drawbacks mentioned in section
\ref{bittorrent}. However, these changes also require some
modifications to the distribution system to support it.

Our solution differs from them in that ...

@@

\section{Conclusion and Future Work}
\label{conclusions}

@@
